---
title: "Meteorite Landings and Near Earth Objects"
author: "Anish Gupta"
output:
  html_document:
   toc: true
   toc_depth: 2
   toc_float:
      collapsed: true
      smooth_scroll: false
editor_options:
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
library(tidyverse)
library(rvest)
library(jsonlite)
library(ggthemes)
library(readxl)
library(lubridate)
library(sf)
library(mapview)
library(plotly)
library(dplyr)
library(maps)
library(packcircles)
library(reshape2)
library(RColorBrewer)
library(countrycode)
library(ggbeeswarm)
library(rworldmap)

knitr::opts_chunk$set(echo = FALSE)
```
# Prepare Data

## Read Data: Meteorite Landings and Near Earth Objects from NASA's website.

For the Meteorite Landings, I initially pulled data using NASA's API. Unfortunately, I found out very late that NASA places an hourly limit of 1,000 requests per hour.
In order to get higher rate limits, one needs to contact NASA.
Due to this reason, I had to use a CSV file to perform my analysis.
I have however, still pulled the 1,000 rows from the API to show some of the data cleaning exercises I performed.
The CSV data is cleaner and easier to work with than the API data.
Most of the work for this project was on data selection, cleaning and preparation.

For Near Earth Objects (NEOs), I used a dataset downloaded from NASA's website showing NEOs 1 year into the future.

```{r read data, message=FALSE, warning=FALSE}
# Read data from all the sources
# Get Meteorite Landings using NASA's API
url <- "https://data.nasa.gov/resource/gh4g-9sfh.json"
meteorite_landings <- url %>% fromJSON()

# Get Meteorite Landings from CSV file downloaded from NASA's website
meteorite_landings_csv <- read_csv("Meteorite_Landings.csv", show_col_types = FALSE)

# Get Near Earth Objects from CSV file downloaded from NASA's website
near_earth_object <- read_csv("NEO Earth Close Approaches.csv", show_col_types = FALSE)

```

## Clean Data

a.  Clean and Wrangle Meteorite Landings for dataset from the API. After cleaning this dataset contains 1000 rows and 14 columns.

```{r clean meteorite landings for dataset from the API, message=FALSE, warning=FALSE}

# Here I split the year on "-" and used map_chr() to get the first part as year. Next, I converted reclong, reclat, and mass to numbers. I set all NA reclong, reclat to 0. Next, I computed the country and continent names based on latitude and longitude.  
clean_meteorite_landings <- meteorite_landings %>%
  mutate(year = str_trim(map_chr(str_split(year, "-"), 1))) %>%
  mutate(reclong = parse_number(reclong)) %>%
  mutate(reclat = parse_number(reclat)) %>%
  mutate(mass = parse_number(mass)) %>%
  mutate(reclong = ifelse(is.na(reclong), 0, reclong),
         reclat = ifelse(is.na(reclat), 0, reclat)) %>%
  mutate(country = map.where(database="world", reclong, reclat)) %>%
  mutate(continent = countrycode(sourcevar = country, origin = "country.name", destination = "continent"))

```

b.  Clean and wrangle Meteorite Landings for dataset from CSV file. The cleaned dataframe contains 45716 rows and 12 columns.

```{r clean meteorite landings for dataset from CSV, message=FALSE, warning=FALSE}

# I renamed `mass (g)` to mass using the rename() function. I set all NA reclong, reclat to 0. Next, I computed the country and continent names based on latitude and longitude. I used the case_when() function to handle special cases for country names and mapped it to the continents.
clean_meteorite_landings_csv <- meteorite_landings_csv %>%
  rename(mass = `mass (g)`) %>%
  mutate(reclong = ifelse(is.na(reclong), 0, reclong),
         reclat = ifelse(is.na(reclat), 0, reclat)) %>%
  mutate(country = map.where(database="world", reclong, reclat)) %>%
  mutate(continent = countrycode(sourcevar = country, origin = "country.name", destination = "continent")) %>%
  mutate(continent = case_when(country == "UK:Northern Ireland" ~ "Europe",
                               country == "Antarctica" ~ "Antarctica",
                               TRUE ~ as.character(continent)))
# Write this dataframe as a CSV file.
write_csv(x = clean_meteorite_landings_csv, "clean_meteorite_landings_csv.csv")
```

c.  Clean and wrangle NEO dataset from CSV file.The cleaned dataframe contains 109 rows and 14 columns.

```{r clean near earth object, message=FALSE, warning=FALSE}

# Here I first split the `Close-Approach (CA) Date` and computed the close_approach_date_time, close_approach_date, and the close_approach_time. I removed the objects (2006 HV5), 7335 (1989 JA), and 349068 (2006 YT13) as their diameter information was not standard. Split and process Diameter field and create new columns called low_diameter, high_diameter, and mean_diameter. Dropped the column `Close-Approach (CA) Date`. I converted the close_approach_date to ymd format.  

clean_near_earth_object <- near_earth_object %>%
  mutate(close_approach_date_time = str_trim(map_chr(str_split(`Close-Approach (CA) Date`,"±"),1))) %>%
  mutate(close_approach_date = str_trim(map_chr(str_split(close_approach_date_time, " "), 1))) %>%
  mutate(close_approach_time = str_trim(map_chr(str_split(close_approach_date_time, " "), 2))) %>%
  filter(Object!= "(2006 HV5)") %>%
  filter(Object!= "7335 (1989 JA)") %>%
  filter(Object!= "349068 (2006 YT13)") %>%
  mutate(low_diameter = str_trim(map_chr(str_split(Diameter, "-"), 1))) %>%
  mutate(high_diameter = str_trim(map_chr(str_split(Diameter, "-"), 2))) %>%
  mutate(low_diameter = str_replace(low_diameter, "m", "")) %>%
  mutate(high_diameter = str_replace(high_diameter, "m", "")) %>%
  mutate(low_diameter = parse_number(low_diameter)) %>%
  mutate(high_diameter = parse_number(high_diameter)) %>%
  mutate(mean_diameter = (low_diameter + high_diameter)/2) %>%
  select(-`Close-Approach (CA) Date`) %>%
  mutate(close_approach_date = ymd(close_approach_date))

# Write dataframe to a CSV file.
write_csv(x = clean_near_earth_object, "clean_near_earth_object.csv")

```

# A study of Meteorite Landings

Meteoroids are known as "space rocks" and their sizes range from dust grains to small asteroids.
Most are pieces of other bodies of rocks that have been broken off.
They can come from comets, asteroids, from the Moon, and other planets.
Some meteoroids can be rocky, or metallic, or a mixture of rock, and metal.
Meteoroids entering the Earth's atmosphere at high speed and burn up, are known as meteors.
They are also referred to as "shooting stars".
When a meteoroid enters the Earth's atmosphere and hits the ground, it is called a meteorite.
When there are lots of meteors in the night sky, there is a meteor shower.

## Top 10 Meteorite Landings by Mass

```{r top 10 meteorite landings csv by Mass, message=FALSE, warning=FALSE}
top_10_meteorite_landings_csv <- clean_meteorite_landings_csv %>%
  select(name, id, nametype, recclass, mass, fall, year, country, reclat, reclong) %>%
  arrange(desc(mass)) %>%
  head(n = 10)
top_10_meteorite_landings_csv %>%
  knitr::kable()
```

## All Known Meteorite Landings in the 21st Century

```{r map recent meteorite landings, message=FALSE, warning=FALSE}
clean_meteorite_landings_csv %>%
  filter(year >= 2000) %>%
  mapview(xcol = "reclong", ycol = "reclat", crs = 4269,cex = "mass",grid = FALSE, legend = FALSE)
```

When you look at this map you can see that most meteorite landings are over land.
This is probably because the ones falling in the oceans are probably rarely recorded.
If you hover over the circles on this map, it will show you the latitude and the longitude information of the locations.
If you click on the circles, it will display all of the details of the particular meteorite.

```{r Craeting a dot plot using the top 100 meteteorite landings, message=FALSE, warning=FALSE, fig.show='hide'}
top_100_meteorite_landings_csv <- clean_meteorite_landings_csv %>%
  arrange(desc(mass)) %>%
  head(n = 100)

top_100_meteorite_landings_csv %>%
  ggplot(aes(year, mass, label = country,colour = fall)) +
  scale_y_continuous(labels = scales::comma) +
  geom_point() +
  ggtitle("Top 100 Meteorite Landings") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45)) +
  xlab("Year") +
  scale_y_continuous(
  name = "Mass in grams",
  trans = "log10",
  labels = scales::comma
  )

ggplotly()
```

This graph shows 100 meteorites with the largest mass.
Meteorites are categorized as "Fell" or "Found".
These categories are shown here in different colors.

## Top 100 Meteorite Landings on a Map.

```{r top 100 meteorite landings of all time on the world map, message=FALSE, warning=FALSE}
top_100_meteorite_landings_csv %>%
  mapview(xcol = "reclong", ycol = "reclat", crs = 4269, grid = FALSE,cex = "mass", zcol = "mass", popup = TRUE, color = "blue",map.types = c("Esri.WorldShadedRelief", "OpenStreetMap.DE"))

```

If you hover over the circles on this map, it will show you the latitude and the longitude information of the locations.
If you click on the circles, it will display all of the details of the particular meteorite.

```{r meteorite landings by country, message=FALSE, warning=FALSE, fig.show='hide'}
meteorite_landings_by_country <- clean_meteorite_landings_csv %>%
  group_by(country) %>%
  summarize(total_landings = n())

```

## Next shown are 20 countries with the most meteorite landings

For this graph, Antarctica, which is listed both as a country and a continent in the cleaned dataset has been excluded as Antarctica is technically not a country.
The country Oman with 3094 landings comes out on the top.
Oman is a country located in southeastern coast of the Arabian Peninsula where the Persian Gulf meets the Arabian Sea.
Oman is a monarchy and has many oil and gas pipelines.

```{r Circle packing chart to show 20 countries with most meteorite landings, message=FALSE, warning=FALSE}
top_20_meteorite_landings_by_country <- meteorite_landings_by_country %>%
  na.omit() %>%
  filter(!country == "Antarctica") %>%
  arrange(desc(total_landings)) %>%
  head(20)

cord_meteorite_landings_by_country <- top_20_meteorite_landings_by_country %>%
  circleProgressiveLayout("total_landings", sizetype = "area")


packing_data_meteorite_landings <- cbind(cord_meteorite_landings_by_country, top_20_meteorite_landings_by_country)

cord_vertices <- circleLayoutVertices(cord_meteorite_landings_by_country)

plot_1 <- ggplot()
plot_1 <- plot_1 + geom_polygon(data = cord_vertices, aes(x, y, group = id, fill = as.factor(id)))
plot_1 <- plot_1 + geom_text(data = packing_data_meteorite_landings, aes(x, y, size = total_landings, label = paste0(country, '\n', total_landings)))
plot_1 <- plot_1 + coord_equal()
plot_1 <- plot_1 + theme_void()
plot_1 <- plot_1 + theme(legend.position = "none")
plot_1 <- plot_1 + ggtitle("Top 20 Meteorite Landings by Country")

nb.cols <- 20
mycolors <- colorRampPalette(brewer.pal(12, "Set3"))(nb.cols)
plot_1 <- plot_1 + scale_fill_manual(values = mycolors)
plot_1 <- plot_1 + theme(plot.title = element_text(hjust = 0.5))
plot_1
```

## A look at the median Meteorite Landings by Continent

As Antarctica does not have any countries, it is not possible to get a true median.
The total number of Meteorite Landings in Antarctica is greater than any other continent at 20,161.
Not including Antarctica, the continent with the highest median landing is Africa at 6.

```{r Meteorite landings by Continent, message=FALSE, warning=FALSE, fig.show='hide'}

clean_meteorite_landings_csv %>%
  filter(!is.na(continent)) %>%
  group_by(continent,country) %>%
  summarize(total_meteorite_by_cont_country = n()) %>%
  ggplot(aes(continent,total_meteorite_by_cont_country, fill = continent)) +
  geom_boxplot() +
  geom_beeswarm() +
  xlab("Continent") +
  scale_y_continuous(
  name = "Meteorite Landings",
  trans = "log10"
  
  ) +
  ggtitle("Meteorite Landings by Continent") +
  theme(plot.title = element_text(hjust = 0.5), legend.position="none") 
 
ggplotly()

```

# A Study of Near Earth Objects

Near Earth Objects (NEOs) are defined as "any small body (comet or asteroid) passing within 1.3 astronomical unit (au) of the Sun".
1 au is the distance from Earth to Sun = \~ 93 million miles.
An NEO is considered to be a Potentially Hazardous Object (PHO) if it passes within 0.05 au of Earth's orbit.
All data in this NEO dataset is less than 0.05 au and therefore, all the objects can be considered to be a PHO.

## Ten Closest NEOs over the Next Year

```{r Closest NEOs over the up coming year, message=FALSE, warning=FALSE}

sorted_clean_near_earth_objects <-clean_near_earth_object %>%
  arrange(`CA DistanceNominal (au)`) %>%
  select(`CA DistanceNominal (au)`, `CA DistanceMinimum (au)`, `V relative(km/s)`, Diameter, mean_diameter
         , close_approach_date)

sorted_clean_near_earth_objects %>%
  head(n = 10) %>%
  knitr::kable()

```

## Mean Diameter of NEOs in a Histogram

This histogram is right skewed.
Most of the mean diameters are between 3 and 117.
The two NEOs with mean diameters of 516, 594, and 606 can be considered to be outliers.

```{r Histogram of mean_diameter, message=FALSE, warning=FALSE, fig.show='hide'}
sorted_clean_near_earth_objects %>%
  ggplot(aes(x = mean_diameter)) +
  geom_histogram( binwidth=3, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Histogram: Mean Diameter of Near Earth Objects") +
  xlab("Mean Diameter") +
  ylab("Count")
 
ggplotly()
  
    
```

## NEOs by Distance from Earth, their Relative Velocity, and Mean Diameter

Here all the NEOs are PHOs as their distance is within 0.05 au.
The NEOs with the smallest distance, largest velocity, and the largest diameters are the most dangerous.

```{r Near Earth Objects by distance, velocity, and diameter, message=FALSE, warning=FALSE, fig.show='hide'}
sorted_clean_near_earth_objects %>%
  ggplot(aes(`CA DistanceNominal (au)`, `V relative(km/s)`, size = mean_diameter, color = mean_diameter, label = close_approach_date)) +
  geom_point() +
  ggtitle("Near Earth Objects") +
  theme(plot.title = element_text(hjust = 0.5),axis.text.x = element_text(angle = 360)) +
  scale_color_continuous(name = "") +
  scale_size(c(1,10))
  
ggplotly()
```

In the graph above the size and color of the points represent the mean diameter.

# Sources

<https://data.nasa.gov/Space-Science/Meteorite-Landings/gh4g-9sfh>

<https://cneos.jpl.nasa.gov/ca/>

<https://www.nsf.gov/attachments/242692/public/JohnsonLindley_PDCONEOOBrief.pdf>

<https://map-rfun.library.duke.edu/01_georeference.html>

<https://www.youtube.com/watch?v=jKIR2-QpaI4>

<https://www.datanovia.com/en/blog/easy-way-to-expand-color-palettes-in-r/>

<https://api.rpubs.com/solomona/494803>

<https://r-graph-gallery.com/220-basic-ggplot2-histogram.html>

<https://www.britannica.com/place/Oman>
